Propositions as Sessions, Mechanically

### 2015-04-14 paper/formalisation.tex
Tweaking some sections.

\textbf{Representation.} As discussed in the previous sections representation
choice is a central issue when formalising programming language
metatheory. Decisions on how to define, for instance, types, calculi and
translations impact on future definitions and properties dependent on
them. Considerations range from how much expressive power provided by the
theorem prover to exploit to the choice of encoding to use for the properties
of interest. Some of these decisions may impact the correspondence between the
formal and informal systems e.g. the empty choice rule and omission of
contraction in CP, and the omission of n-ary branch and choice from GV, and
such things are necessarily a tradeoff between simplicity and
adequacy.

@@

\textbf{Library reuse.} One of the design goals of the development was to
reuse existing libraries wherever possible. Two such instances proved
extremely useful, first is the Metatheory library as already
mentioned. Second, is the library of tactics provided by the Software
Foundations online book~\cite{SF}. The tactics helped in many ways including
simplifying the introduction of assertions and providing convenient shorthands
for commonly executed sequences of tactics. An example for simplifying
assertions would be the \coqe!forwards! tactic permitting lemma
instantiation. Consider the following example (during the proof of $assoc_1$):
\begin{tabular}{l}
\begin{coq}
1 subgoal
(* ... other elements of the proof context ... *)
PER : Permutation (x $\sim$ B ++ $\GD$P) ($\GD$P0++$\GD$Q0)
UN : uniq (x $\sim$ B ++ $\GD$P)
\end{coq}\\ \hline
\begin{coq}
$\nu$ A.(P $\mid$ $\nu$ B.($\open{0}{1}{Q}$ $\mid$ R)) $\tpcp$ $\Gamma$
\end{coq}
\end{tabular}
\\~\\

In order to proceed with proving the goal, it is necessary to show that
$\GD$P0++$\GD$Q0 is unique. The proof follows from \key{PER} and \key{UN} upon
application of an appropriate lemma, and there are two possible ways to prove
this using plain Coq. One way is to apply the lemma to one of \key{PER} or
\key{UN} thereby replacing the chosen hypothesis with the conclusion of the
lemma (termed \textit{forward reasoning}). The opposite of forward reasoning,
namely backward reasoning, does not apply since the goal has nothing to do
uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert! tactic one
can utilise backward reasoning. An assertion temporarily focuses a different
goal (the asserted goal) which may then be tackled using tactics in the
standard way. Unfortunately, this approach requires the explicit writing of
the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script of the
lemma cluttering the proof script sensitive to change because \coqe!$\GD$P0!
and \coqe!$\GD$Q0! are ephemeral names introduced by prover.

The \coqe!forwards! tactic avoids the problem of explicit names by allowing
the uniqueness property to be proven succiently:
\begin{coq}
forwards UN1: uniq_perm PER UN.
\end{coq}

>>>>

\textbf{Representation.} As discussed in the previous sections representation
choice is a central issue when formalising programming language
metatheory. Decisions on how to define, for instance, types, calculi and
translations impact on future definitions and properties dependent on
them. Considerations range from how much expressive power provided by the
theorem prover to exploit to the choice of which encoding to use for the
properties of interest. Some of these decisions may impact the correspondence
between the formal and informal systems e.g. the empty choice rule and
omission of contraction in CP, and the omission of $n$-ary branch and choice
from GV, and such things are necessarily a tradeoff between simplicity and
adequacy.

@@

\textbf{Library reuse.} One of the design goals of the development was to
reuse existing libraries wherever possible. Two such instances proved
extremely useful, first is the Metatheory library as already
mentioned. Second, is the library of tactics provided by the Software
Foundations online book~\cite{SF}. The tactics helped in many ways including
simplifying the introduction of assertions and providing convenient shorthands
for commonly executed sequences of tactics. An example for simplifying
assertions would be the \coqe!forwards! tactic permitting lemma
instantiation. Consider the following example (during the proof of
\coqe!assoc_1!):
\begin{tabular}{l}
\begin{coq}
1 subgoal
(* ... other elements of the proof context ... *)
PER : Permutation (x $\sim$ B ++ $\GD$P) ($\GD$P0++$\GD$Q0)
UN : uniq (x $\sim$ B ++ $\GD$P)
\end{coq}\\ \hline
\begin{coq}
$\nu$ A.(P $\mid$ $\nu$ B.($\open{0}{1}{Q}$ $\mid$ R)) $\tpcp$ $\Gamma$
\end{coq}
\end{tabular}
\\~\\

In order to proceed with proving the goal, it is necessary to show that
$\GD$P0++$\GD$Q0 is unique. The proof follows from \key{PER} and \key{UN} upon
application of an appropriate lemma and there are two possible ways to prove
this using plain Coq. One way is to apply the lemma to one of \key{PER} or
\key{UN} thereby replacing the chosen hypothesis with the conclusion of the
lemma (termed \textit{forward reasoning}). The opposite of forward reasoning,
namely backward reasoning, does not apply since the goal has nothing to do
with the uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert!
tactic one can utilise backward reasoning. An assertion temporarily focuses a
different goal (the asserted goal) which may then be tackled using tactics in
the standard way. Unfortunately, this approach requires the explicit writing
of the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script
cluttering it with unnecessary details and making it sensitive to change
because \coqe!$\GD$P0! and \coqe!$\GD$Q0! are ephemeral names introduced by
theorem prover.

The \coqe!forwards! tactic avoids the problem of explicit names by allowing
the uniqueness property to be proven succinctly:
\begin{coq}
forwards UN1: uniq_perm PER UN.
\end{coq}

### 2015-04-14 paper/futurework.tex
Tweaking some sections.

\textbf{CP-GV system.} The issues regarding cut elimination suggest there
should be greater study into calculi based on the sequent calculus. It is
unclear to what extent process calculi such as CP and lambda calculi can be
unified to use the same framework or library. The different considerations of
sequent and process calculi perhaps mandate a different library or
framework. It would be interesting to consider CP in the setting described by
\citeauthor{Tews:2013,Dawson:2010,Dawson:2002}, where explicit representation
is given to derivations. The aim here would be to make progress in proving the
cut elimination theorem for CP, but it is unclear how linearity and
permutation of environments would be handled in this setting. Otherwise,
defining an appropriate termination measure for the current version of CP may
be possible, though it remains unclear how to obtain the desired relationship
between subprocesses of cut (as explained in \S~\ref{sec:cutelim}).

Further progress with the graph relation version of the translation could be
made in the current setting aiming to limit the duplication discussed
previously The translation would be substantially affected by the
aforementioned change to CP because the derivations in the conlusions of the
rules would have to be constructed. In this case, it may be worth
investigating representing derivations in GV explicitly.

\citeauthor{Lindley:2014:SAP} extend \citeauthor{Wadler:2014}'s work to
support polymorphism in GV and n-ary branch and choice in CP. Future work is
to support these constructs.

\textbf{Beyond Linear.} Most session-based type systems treat channels as
linear resources which may not be aliased. The ``adoption and focus''
system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
resources. The type system requires special language features, e.g. to manage
capabilities. \citeauthor{Fahndrich:2002} do not provide operational semantics
or soundness results for the type system. It would be interesting to
investigate these properties formally in a session type setting. The ``focus''
rule (and linearity generally) enforces an invariant on the focussed object
(channel) such that the type is maintained during the focus operation. Such a
rule would appear to only benefit recursive session types. Indeed, focussing
on a non-recursive channel would appear pointless because one is prevented
from communicating across it for the entire scope of the focus. Moreover, it
would be undesirable to permit modification of a session type without
guaranteeing exclusive access to the channel since this could lead to invalid
communication.

>>>>

\textbf{CP-GV system.} The issues regarding cut elimination suggest there
should be greater study of calculi based on the sequent calculus. It is
unclear to what extent process calculi such as CP and $\lambda$ calculi can be
unified to use the same framework or library. The different considerations of
sequent and process calculi perhaps mandate a different approach for each. It
would be interesting to consider CP in the setting described by
\citeauthor{Tews:2013} and \citeauthor{Dawson:2010}, where explicit
representation is given to derivations. The aim here would be to make progress
on proving the cut elimination theorem for CP, but it is unclear how linearity
and permutation of environments would be handled in this setting. Otherwise,
defining an appropriate termination measure for the current version of CP may
be possible, though it remains unclear how to obtain the desired relationship
between subprocesses of cut (as explained in \S~\ref{sec:cutelim}).

Further progress with the graph relation version of the translation could be
made in the current setting aiming to limit the duplication discussed
previously. The translation would be substantially affected by the
aforementioned change to CP because the derivations in the conlusions of the
rules would have to be constructed. In this case, it may be worth
investigating representing derivations in GV explicitly.

\citeauthor{Lindley:2014:SAP} extend \citeauthor{Wadler:2014}'s work to
support polymorphism in GV and $n$-ary branch and choice in CP. Future work is
to support these constructs.

\textbf{Beyond Linear.} Most session-based type systems treat channels as
linear resources which may not be aliased. The ``adoption and focus''
system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
resources.\todo{say what adopt and focus do} The type system requires special
language features, e.g. to manage capabilities. \citeauthor{Fahndrich:2002} do
not provide operational semantics or soundness results for the type system. It
would be interesting to investigate these properties formally in a session
type setting. The ``focus'' rule enforces an invariant on the focussed object
(channel) such that the type is maintained during the focus operation. The
object may be aliased while it is under a focus operation. Such a rule would
appear to only benefit recursive session types. Indeed, focussing on a
non-recursive channel would appear pointless because one is prevented from
communicating across it for the entire scope of the focus. Moreover, it would
be undesirable to permit modification of a session type without guaranteeing
exclusive access to the channel since this could lead to invalid
communication.
